Large-scale harmonic patterns clustering

### 2010-03-10 Paper/Paper.tex
improved

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{code}
\end{center}
\caption{{A typical code from a code book of size $200$. Patterns represented
$2$ bars and the pattern length was set to $16$.}}
\label{fig:code}
\end{figure}

@@

uspop2002 serves as testing set to measure how well a code book learned on
the Cowbell data set can represent new songs. We get Echo Nest features
for $8651$ songs from that dataset.

\section{Algorithm}\label{sec:algo}
In this section we present the creation of the harmonic patterns from
the Echo Nest data. Then, we explain the online vector quantization algorithm
used to learn a code book of those patterns.

@@

\section{Experiments}\label{sec:experiments}
In this section we present general experiments that describe our method.
We start by giving some details on how we selected our training set and
how to measure error. Secondly, we experiment with different pattern sizes.
Thirdly, we look at the effect of the size of the
data set on the training. Then, we show that the patterns are not random,
which is expected from music data. Finally, we show examples of patterns,
code books and clusters.

@@

\subsection{Data Size}
We are interesting to see how the amount of training data changes
the quality of our code.
We experiment with different data sizes for a fixed size of codebook and a 
fixed code length. The data points are chose randomly from the whole
data set.
Code length: 1 bar resized to size 4. Code book size: 100
We measure the distortion on our test set. We hope for some convergence.
See results in Figure \ref{fig:sizes}.

@@

$500$ samples per code in the code book seems a good data size, see
Figure \ref{fig:sizes} for size $50$K.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.99\columnwidth]{data_sizes}
\end{center}
\caption{{Distortion for a code book of size $100$ encoding one bar
at a time, a bar being represented by $4$ chroma feature vector.
Therefore, a code or a pattern is of size $12$ x $4$ = $48$.
Distortion measure on the test data set. Data size tested range
from $10$ thousands to $2$ millions. Patterns were selected at
random from the dataset of approximately $3,7$ million patterns.}}
\label{fig:sizes}
\end{figure}

\iffalse
\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
size & $5$K & $10$K & $50$K & $100$K & $500$K & $2$M  \\ \hline
distortion & $$ & $0.143$ & $0.142$ & $0.142$ & $0.142$ & $$\\ \hline
\end{tabular}
\end{center}
\caption{{Distortion for a code book of size $100$ encoding one bar
at a time, a bar being represented by $4$ chroma feature vector.
Therefore, a code or a pattern is of size $12$ x $4$ = $48$.
Distortion measure on the test data set. Data size tested range
from $10$ thousands to $2$ millions. Patterns were selected at
random from the dataset of approximately $3,7$ million patterns.}}
\label{tab:datasize}
\end{table}
\fi
% RESULTS:    
% 1bar s4     avg_distOLD    avg_dist
% 1K          0.150370
% 5K          0.145327
% 10K         0.143567
% 50K         0.142257
% 100K        0.142195
% 250K        0.141796
% 500K        0.142081
% 1M          0.142118
% 2M          0.141596
% 1bar s4 first X samples
% 1K          0.157142
% 5K          0.150045
% 10K         0.148357
% 50K         0.146069
% 100K        0.145765
% 250K        0.145819
% 500K        0.145772
% 1M          0.145924
% 2M          0.146065
% 0bar s4 (and test set same, we could do 1bar s4 in test set...)
% 1K          0.155474
% 5K          0.149619
% 10K         0.148056
% 50K         0.146328
% 100K        0.146004
% 250K        0.146024
% 500K        0.145962
% 1M          0.145873
% 2M          0.145955
% 2bar s8     
% 1K          0.171300
% 5K          0.165960
% 10K         0.165063
% 50K         0.162439
% 100K        0.161588
% 250K        0.161910
% 500K        0.161233
% 1M          0.161057
% 2M          0.161068

\subsection{Codebook Size}
Results in Table \ref{tab:cbsize}.

\begin{table}
\begin{center}
\begin{tabular}{|l|l|c|}
\hline
\# codes & \# samples & avg. dist \\ \hline \hline
1 & $500$ & $0.066079$ \\
10 & $5$K & $0.045690$ \\
50 & $25$K & $0.038456$ \\
100 & $50$K & $0.036725$ \\
200 & $100$K & $0.033923$ \\
500 & $250$K & $$ \\
1000 & $500$K & \\ \hline
\end{tabular}
\end{center}
\caption{{Code book size.}}
\label{tab:cbsize}
\end{table}
% RESULTS:      avg_distsOLD    avg_dist
% code 1        0.195939        0.066079
% code 10       0.161812        0.045690
% code 50       0.147472        0.038456
% code 100      0.142194        0.036725
% code 200      0.137548        0.033923
% code 500      0.131211        0.

@@

See code book size experiments in Table \ref{tab:cbsize}.

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|}
\hline
\# codes & \# bars & avg. dist \\ \hline \hline
2 & $1$ & $0.058435$ \\
4 & $2$ & $0.054792$ \\
16 & $4$ & $0.055489$ \\
256 & $8$ & $0.054860$ \\ \hline
2 & $2$ beats* & $0.066416$ \\
4 & $1$ & $0.050712$ \\
16 & $2$ & $0.050172$ \\
256 & $4$ & $0.049894$ \\ \hline
\end{tabular}
\end{center}
\caption{{Code book size in function of the number of bars to encode.
All experiments done wtih $256 \times 500 = 128$K samples randomly chosen.
* For the $2$ beats experiment, we take the first half of every beats.
It is different from ignoring bars and taking every $2$ beats. It is more
coherent with the other settings ($1-2-4-8$ bars).}}
\label{tab:size_pattern}
\end{table}
% RESULTS:           avg_distsOLD   avg_dist
%2 codes 1 bar       0.180750       0.058435
%4 codes 2 bars      0.183631       0.054792
%16 codes 4 bars     0.184078       0.055489
%256 codes 8 bars    0.183464       0.054860
%---------------
%2 codes 2 beats     0.166840       0.046955
%2 codes 2 beats                    0.066416
% based on bars
%4 codes 1 bar       0.171495       0.050712
%16 codes 2 bars     0.173767       0.050172
%256 codes 4 bars    0.174169       0.049894

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{{LLE visualization of the code book.}}
\label{fig:lle}
\end{figure}

Another way is simply to present the codes in order of cluster size.
We encode the training or testing set as explained in Subsection 
\ref{ssec:setting} and we count how many patterns were closer to a particular
code than any other code. In Figure \ref{fig:firstcodes}, we show the $25$
most used codes out of a code book of size $200$.
Code from Figure \ref{fig:code} is the 37th most used code in that
code book. In Figure \ref{fig:closep}, we show a sample of patterns that
fall into its cluster.

@@

\subsection{Training on our Codebook}
NMF provides us with a way of measuring how much information is retained
in our code book. More specifically, we can train a code book of size $500$
using our online VQ algorithm, then decompose that code into $10$ basis
using NMF. We can compare that to $10$ basis from training NMF on the
whole data.

>>>>

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{code}
\end{center}
\caption{{A typical code from a codebook of size $200$. Patterns represented
$2$ bars and the pattern length was set to $16$.}}
\label{fig:code}
\end{figure}

@@

uspop2002 serves as testing set to measure how well a codebook learned on
the Cowbell data set can represent new songs. We get Echo Nest features
for $8651$ songs from that dataset.

\section{Algorithm}\label{sec:algo}
In this section we present the creation of the harmonic patterns from
the Echo Nest data. Then, we explain the online vector quantization algorithm
used to learn a codebook of those patterns.

@@

\section{Experiments}\label{sec:experiments}
In this section we present general experiments that describe our method.
We start by giving some details on how we selected our training set and
how to measure error. Secondly, we experiment with different pattern sizes.
Thirdly, we look at the effect of the size of the
data set on the training. The next experiment look at the size of the
codebook. Then, we show that the patterns are not random,
which is expected from music data. Finally, we show examples of patterns,
codebooks and clusters.

@@

\subsection{Data Size}
We are interesting to see how the amount of training data changes
the quality of our code.
We experiment with different data sizes for a fixed size of codebook and a 
fixed code length. The data points are chose randomly from the whole
data set.
Code length: 1 bar resized to size 4. Codebook size: 100
We measure the distortion on our test set. We hope for some convergence.
See results in Figure \ref{fig:sizes}.

@@

$500$ samples per code in the codebook seems a good data size, see
Figure \ref{fig:sizes} for size $50$K.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.99\columnwidth]{data_sizes}
\end{center}
\caption{{Distortion for a codebook of size $100$ encoding one bar
at a time, a bar being represented by $4$ chroma feature vector.
Therefore, a code or a pattern is of size $12$ x $4$ = $48$.
Distortion measure on the test data set. Data size tested range
from $10$ thousands to $2$ millions. Patterns were selected at
random from the dataset of approximately $3,7$ million patterns.}}
\label{fig:sizes}
\end{figure}

\iffalse
\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
size & $5$K & $10$K & $50$K & $100$K & $500$K & $2$M  \\ \hline
distortion & $$ & $0.143$ & $0.142$ & $0.142$ & $0.142$ & $$\\ \hline
\end{tabular}
\end{center}
\caption{{Distortion for a codebook of size $100$ encoding one bar
at a time, a bar being represented by $4$ chroma feature vector.
Therefore, a code or a pattern is of size $12$ x $4$ = $48$.
Distortion measure on the test data set. Data size tested range
from $10$ thousands to $2$ millions. Patterns were selected at
random from the dataset of approximately $3,7$ million patterns.}}
\label{tab:datasize}
\end{table}
\fi
% RESULTS:    
% 1bar s4     avg_distOLD    avg_dist
% 1K          0.150370
% 5K          0.145327
% 10K         0.143567
% 50K         0.142257
% 100K        0.142195
% 250K        0.141796
% 500K        0.142081
% 1M          0.142118
% 2M          0.141596
% 1bar s4 first X samples
% 1K          0.157142
% 5K          0.150045
% 10K         0.148357
% 50K         0.146069
% 100K        0.145765
% 250K        0.145819
% 500K        0.145772
% 1M          0.145924
% 2M          0.146065
% 0bar s4 (and test set same, we could do 1bar s4 in test set...)
% 1K          0.155474
% 5K          0.149619
% 10K         0.148056
% 50K         0.146328
% 100K        0.146004
% 250K        0.146024
% 500K        0.145962
% 1M          0.145873
% 2M          0.145955
% 2bar s8     
% 1K          0.171300
% 5K          0.165960
% 10K         0.165063
% 50K         0.162439
% 100K        0.161588
% 250K        0.161910
% 500K        0.161233
% 1M          0.161057
% 2M          0.161068

\subsection{Codebook Size}
We experiment with the size of the codebook and its affect on the
performance of the encoding. Results are shown in Table \ref{tab:cbsize}.
Note that we trained on a number of samples depending on the codebook
size, i.e. we used $500$ samples per code to learn. This is consistent
with our findings in Figure \ref{fig:sizes}.

\begin{table}
\begin{center}
\begin{tabular}{|l|l|c|}
\hline
\# codes & \# samples & avg. dist \\ \hline \hline
1 & $500$ & $0.066079$ \\
10 & $5$K & $0.045690$ \\
50 & $25$K & $0.038456$ \\
100 & $50$K & $0.036725$ \\
200 & $100$K & $0.033923$ \\
500 & $250$K & $$ \\
1000 & $500$K & \\ \hline
\end{tabular}
\end{center}
\caption{{Codebook size.}}
\label{tab:cbsize}
\end{table}
% RESULTS:      avg_distsOLD    avg_dist
% code 1        0.195939        0.066079
% code 10       0.161812        0.045690
% code 50       0.147472        0.038456
% code 100      0.142194        0.036725
% code 200      0.137548        0.033923
% code 500      0.131211        0.

@@

See codebook size experiments in Table \ref{tab:cbsize}.

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|}
\hline
\# codes & \# bars & avg. dist \\ \hline \hline
2 & $1$ & $0.058435$ \\
4 & $2$ & $0.054792$ \\
16 & $4$ & $0.055489$ \\
256 & $8$ & $0.054860$ \\ \hline
2 & $2$ beats* & $0.066416$ \\
4 & $1$ & $0.050712$ \\
16 & $2$ & $0.050172$ \\
256 & $4$ & $0.049894$ \\ \hline
\end{tabular}
\end{center}
\caption{{Codebook size in function of the number of bars to encode.
All experiments done wtih $256 \times 500 = 128$K samples randomly chosen.
* For the $2$ beats experiment, we take the first half of every beats.
It is different from ignoring bars and taking every $2$ beats. It is more
coherent with the other settings ($1-2-4-8$ bars).}}
\label{tab:size_pattern}
\end{table}
% RESULTS:           avg_distsOLD   avg_dist
%2 codes 1 bar       0.180750       0.058435
%4 codes 2 bars      0.183631       0.054792
%16 codes 4 bars     0.184078       0.055489
%256 codes 8 bars    0.183464       0.054860
%---------------
%2 codes 2 beats     0.166840       0.046955
%2 codes 2 beats                    0.066416
% based on bars
%4 codes 1 bar       0.171495       0.050712
%16 codes 2 bars     0.173767       0.050172
%256 codes 4 bars    0.174169       0.049894

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{{LLE visualization of the codebook.}}
\label{fig:lle}
\end{figure}

Another way is simply to present the codes in order of cluster size.
We encode the training or testing set as explained in Subsection 
\ref{ssec:setting} and we count how many patterns were closer to a particular
code than any other code. In Figure \ref{fig:firstcodes}, we show the $25$
most used codes out of a codebook of size $200$.
Code from Figure \ref{fig:code} is the 37th most used code in that
codebook. In Figure \ref{fig:closep}, we show a sample of patterns that
fall into its cluster.

@@

\subsection{Training on our Codebook}
NMF provides us with a way of measuring how much information is retained
in our codebook. More specifically, we can train a codebook of size $500$
using our online VQ algorithm, then decompose that code into $10$ basis
using NMF. We can compare that to $10$ basis from training NMF on the
whole data.
